Legacy Data Confounds Genomics Studies

### 2019-03-06 legacy.tex
minor updates

We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
Over 3826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with population as a covariate to account for population structure.
This method identifies a total of 11,416 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.001$ \citep{Benjamini2006}.

@@

\subsection{Cell line or technical artifact}

To determine whether this signal is the result of a cell line or technical artifact, we investigated other sequencing projects that used the same cell lines.
Conveniently, the HapMap project sequenced some of the same cell lines as in the 1kGP with a different sequencing platform.
The technological artifacts in one platform aren't expected to be present in another sequencing platform as the technological biases vary from platform to platform \citep{Minoche2011}.
If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
From the list of 11,416 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
This result suggests that there is no cell line artifact and that the variants identified in the 1kGP are the result of technological artifact.

In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
After restricting the analysis to individuals in both datasets, we find that 9 out of the 296 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
While this is more than 0.1\% of the sites, we suspect that they are indeed false associations to $Q$.

@@

One of these studies used the 1kGP in their reference panel for imputation \citep{tian2017genome}.
Two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
And one study used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.

>>>>

We therefore performed the same reverse GWAS approach in all populations independently, and similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected, with on average four times as many significantly associated sites compared to the phase 3 populations.
Over 3,826 variants were independently associated to low $Q$ in at least two populations with $ p < 10^{-6}$ in each (Figure \ref{OverLap}).

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with population as a covariate to account for population structure.
This method identifies a total of 11,416 statistically significant variants associated to $Q$ (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). 
We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.

@@

\subsection{Cell line or technical artifact}
\luke{Do we keep HapMap? It may be worth keeping some of it to explain why 2/4 papers impute using HapMap and find the same variants.}
To determine whether this signal is the result of a cell line or technical artifact, we investigated other sequencing projects that used the same cell lines.
Conveniently, the HapMap project sequenced some of the same cell lines as in the 1kGP with a different sequencing platform.
The technological artifacts in one platform aren't expected to be present in another sequencing platform as the technological biases vary from platform to platform \citep{Minoche2011}.
If the variants we've identified are the result of a cell line artifact, then we would expect the same variants to be present in the data from two separate platforms.
From the list of 11,416 variants identified as being associated to $Q$, only 17 variants were present in HapMap \citep{HapMap2005}.
Indeed, we find that while there are some variants that are present in HapMap, the number is within the expected number of false discoveries based on our FDR adjustment.
This result suggests that there is no cell line artifact and that the variants identified in the 1kGP are the result of technological artifact.

In 2017, Lan et al. resequenced 83 Han Chinese individuals from the 1kGP \citep{Lan2017}.
After restricting the analysis to individuals in both datasets, we find that 9 out of the 296 the suspicious variants from the single population test are present in the resequenced data \ref{90HanSFS}.
While this is more than 0.1\% of the sites, we suspect that some of these variants may be associated to $Q$ as a result of incorrect genotyping.

@@

One of these studies used the 1kGP in their reference panel for imputation \citep{tian2017genome} and another used the 1kGP sequence data and cell lines directly \citep{Mandage2017}.
Two studies genotyped individuals and imputed the data using the HapMap II as a reference  database for imputation \citep{Kraja2011, Ebejer2013}.
